Classless Java: Tuning Java Interfaces

### 2015-12-09 paper/sections/Overview.tex
S2 small fixes

\noindent There are some annoyances here. Firstly, the sole
purpose of the classes is to provide a way to instantiate
objects. Although (in this case) it takes only one line of code to
provide each of those classes, this code is essentially boilerplate
code, which does not add behaviour to the system. Secondly,
the namespace gets filled with three additional types. For example,
both \texttt{Horse} and \texttt{HorseImpl} are needed: \texttt{Horse}
is needed because it needs to be an interface so that \texttt{Pegasus}
can use multiple inheritance; and \texttt{HorseImpl} is needed to
provide object instantiation.
Note that, for this very simple animal system, plain Java 8 anonymous
classes can be used to avoid these problems.  We could have simply
instantiated Pegasus using:

@@

\noindent The effect of the annotations is that a static \emph{factory} method called
\texttt{of} is automatically added to the interfaces. With the
\texttt{of} method a Pegasus object is instantiated as follows:

@@

\noindent The \texttt{of} method provides an alternative to a
constructor, which is missing from interfaces. The following code,
shows the code corresponding to the \texttt{Pegasus} interface
after the \mixinAnn{} annotation is processed:

@@

\noindent Note that the generated code is transparent to a user, which
only sees the original code with the \mixin annotation. Compared to the pure
Java solution in Section~\ref{sec:simple}, the solution using object interfaces
has the advantage of providing a direct mechanism for object
instantiation, which avoids adding boilerplate classes to the
namespace.

@@

The animal system modelled so far is a simplified version of the
system presented in the left-side of Figure~\ref{fig:pegasus}, which still does not appear to
justify the \mixinAnn{} annotation.  Now moving on to modelling
the complete animal system where \Q@Animal@s will include a \Q@location@
representing their position in space, we shall
see that modelling stateful components using plain Java 8 quickly becomes
cumbersome. The \mixinAnn{} annotation comes to rescue here and
allows avoiding significant amounts of boilerplate code.

@@

\vspace{-5pt}
\paragraph{\Q@Point2D@ with Object Interfaces}
This obvious ``constructor'' code can be automatically generated by our \mixin
annotation.  By annotating the interface \Q@Point2D@, a variation of the shown
static method \texttt{of} will be generated, mimicking the functionality of a
simple-minded constructor. \mixin first looks at the abstract methods and detects
what are the fields, then generates an \Q@of@ method with one parameter for each
of them. That is, we can just write

@@

\paragraph{\texttt{wither} methods in Object Interfaces}
The pattern of creating a new object by reusing most information from an old
object is very common when programming with immutable
data-structures. As such, it is
supported in our code generation as \Q@with@ methods. For example:
\begin{lstlisting}
@Obj interface Point2D {
  int x();    int y(); // getters
  Point2D withX(int val);    Point2D withY(int val); /* with methods*/ }
\end{lstlisting}

\noindent Using \texttt{with} methods, the point \texttt{p} can be created instead using:

@@

\paragraph{Summary} Dealing with state (mutable or not) in object interfaces
relies on a notion of abstract state, where only methods that interact with
state are available to users. Object interfaces provide support for four
different types of abstract state operations, which are summarized in
Figure~\ref{fig:abstractstate}.  The abstract state operations are determined by
naming conventions and the types of the methods. Fluent setters are a variant of
conventional setters, and are discussed in detail in Section~\ref{sec:dsls}.

@@

\paragraph{Unsatisfactory class-based solutions to field type refinement}
In Java if we define an animal class with a field we have a set of
unsatisfactory options in front of us:
\begin{itemize}
\item Define a \Q@Point3D@ field in \Q@Animal@: this is bad since all animals
  would require more than needed.
  %Also it requires the programmer to predict the future, or
  Also it requires adapting the old code to accommodate for new evolutions.

@@

We may be tempted to assume that a language extension is needed.
%Instead, with object interfaces, another approach is possible
Instead, the \emph{restriction} of (object) interfaces to have no
fields, enlighten us that another approach is possible; often in programming languages ``freedom is slavery''.

\paragraph{Field type refinement with object interfaces}
Object interfaces address the challenge of type-refinement as follows:
\begin{itemize}
\item by \emph{covariant method overriding} the return type of
  \texttt{location()} is refined to \texttt{Point3D};
\item by \emph{overloading} a new setter for location is defined with a more
  precise type;
\item a \Q@default@ setter implementation with the old signature is provided.
\end{itemize}

@@

\noindent \Q@Point3D@ includes a
\Q@with@ method, taking a \Q@Point2D@ as an argument.
Other wither methods (such as \Q@withX@) functionally update a field at a time.  This can be
inefficient, and sometimes hard to maintain.  Often we want to update multiple
fields simultaneously, for example using another object as source.  Following
this idea, the method \Q@with(Point2D)@ is an example of a (functional)
properties updater: it takes a certain type of object and returns a copy of the
current object where all the fields that match fields in the parameter
object are updated to the corresponding value in the parameter. The idea is that
the result should be like \Q@this@, but modified to be as similar as possible to the parameter.

With the new \Q@with@ method we may use the information for
\Q@z@ already stored in the object to forge an appropriate \Q@Point3D@
to store. Note how all the information about what fields sit in
\Q@Point3D@ and in \Q@Point2D@ is properly encapsulated in the
\Q@with@ method, and is transparent to the implementer of \Q@Bird@.

@@

\paragraph{Generated boilerplate}
Just to give a feeling on how much boring code \mixin is generating, we show the
generated code for the \texttt{Point3D} in Figure~\ref{fig:boilerplate}. The
generated code is mechanical. Writing such code by hand is error-prone. For
example a distracted programmer may swap the arguments of calls to
\Q@Point3D.of@.  Note how \Q@with-@ methods are automatically refined in their
return type, so that code like:

@@

\subsection{Advanced Multiple Inheritance with Object Interfaces}
Finally, defining \texttt{Pegasus} is as simple as we did in the simplified
(and stateless) version on the right-side of Figure~\ref{fig:pegasus}.
 Note how even the non-trivial pattern for field type refinement is
transparently composed, and \texttt{Pegasus} has a \Q@Point3D@
\Q@location@.%  This works because \Q@Horse@ do not perform
% any field type refinement, otherwise we may have to choose/create a
% common subtype in order for \texttt{Pegasus} to exists.

>>>>

\noindent There are some annoyances here. Firstly, the sole
purpose of the classes is to provide a way to instantiate
objects. Although (in this case) it takes only one line of code to
provide each of those classes, this code is essentially boilerplate
code, which does not add behaviour to the system. Secondly,
the namespace gets filled with three additional types. For example,
both \texttt{Horse} and \texttt{HorseImpl} are needed: \texttt{Horse}
is needed because it needs to be an interface so that \texttt{Pegasus}
can use multiple inheritance; and \texttt{HorseImpl} is needed to
provide object instantiation.
Note that, for this very simple animal system, plain Java 8 anonymous
classes can be used to avoid these problems.  We could have simply
instantiated \texttt{Pegasus} using:

@@

\noindent The effect of the annotations is that a static \emph{factory} method called
\texttt{of} is automatically added to the interfaces. With the
\texttt{of} method a \texttt{Pegasus} object is instantiated as follows:

@@

\noindent The \texttt{of} method provides an alternative to a
constructor, which is missing from interfaces. The following code 
shows the code corresponding to the \texttt{Pegasus} interface
after the \mixinAnn{} annotation is processed:

@@

\noindent Note that the generated code is not transparent to a user, who 
only sees the original code with the \mixin annotation. Compared to the pure
Java solution in Section~\ref{sec:simple}, the solution using object interfaces
has the advantage of providing a direct mechanism for object
instantiation, which avoids adding boilerplate classes to the
namespace.

@@

The animal system modelled so far is a simplified version of the
system presented in the left-side of Figure~\ref{fig:pegasus}, which still does not appear to
justify the \mixinAnn{} annotation.  Now moving on to modelling
the complete animal system where \Q@Animal@s will include a \Q@location@
representing their positions in space, we shall
see that modelling stateful components using plain Java 8 quickly becomes
cumbersome. The \mixinAnn{} annotation comes to rescue here and
 avoids significant amounts of boilerplate code.

@@

\vspace{-5pt}
\paragraph{\Q@Point2D@ with object interfaces}
This obvious ``constructor'' code can be automatically generated by our \mixin
annotation.  By annotating the interface \Q@Point2D@, a variation of the shown
static method \texttt{of} will be generated, mimicking the functionality of a
simple-minded constructor. \mixin first looks at the abstract methods and detects
what are the fields, then generates an \Q@of@ method with one parameter for each
of them. That is, we can just write

@@

\paragraph{\texttt{with-} methods in object interfaces}
The pattern of creating a new object by reusing most information from an old
object is very common when programming with immutable
data-structures. As such, it is
supported in our code generation as \Q@with-@ methods. For example:
\begin{lstlisting}
@Obj interface Point2D {
  int x();    int y(); // getters
  Point2D withX(int val);    Point2D withY(int val); /* with methods*/ }
\end{lstlisting}

\noindent Using \texttt{with-} methods, the point \texttt{p} can also be created
by:

@@

\paragraph{Summary} Dealing with state (mutable or not) in object interfaces
relies on a notion of abstract state, where only \emph{methods} that interact with
state are available to users. Object interfaces provide support for four
different types of abstract state operations, which are summarized in
Figure~\ref{fig:abstractstate}.  The abstract state operations are determined by
naming conventions and the types of the methods. Fluent setters are a variant of
conventional setters, and are discussed in detail in Section~\ref{sec:dsls}.

@@

\paragraph{Unsatisfactory class-based solutions to field type refinement}
In Java if we want to define an animal class with a field we have a set of
unsatisfactory options in front of us:
\begin{itemize}
\item Define a \Q@Point3D@ field in \Q@Animal@: this is bad since all animals
  would require more than needed.
  %Also it requires the programmer to predict the future, or
  Also it requires adapting the old code to accommodate for new evolutions.

@@

We may be tempted to assume that a language extension is needed.
%Instead, with object interfaces, another approach is possible
Instead, the \emph{restriction} of (object) interfaces to have no
fields enlightens us that another approach is possible; often in programming languages ``freedom is slavery''.

\paragraph{Field type refinement with object interfaces}
Object interfaces address the challenge of type-refinement as follows:
\begin{itemize}
\item by \emph{covariant method overriding}, the return type of
  \texttt{location()} is refined to \texttt{Point3D};
\item by \emph{overloading}, a new setter for location is defined with a more
  precise type;
\item a \Q@default@ setter implementation with the old signature is provided.
\end{itemize}

@@

\noindent \Q@Point3D@ includes a
\Q@with@ method, taking a \Q@Point2D@ as an argument.
Other wither methods (such as \Q@withX@) functionally update a field one at a time.  This can be
inefficient, and sometimes hard to maintain.  Often we want to update multiple
fields simultaneously, for example using another object as source.  Following
this idea, the method \Q@with(Point2D)@ is an example of a (functional)
properties updater: it takes a certain type of object and returns a copy of the
current object where all the fields that match fields in the parameter
object are updated to the corresponding value in the parameter. The idea is that
the result should be like \Q@this@, but modified to be as similar as possible to the parameter.

With the new \Q@with@ method we may use the information for
\Q@z@ already stored in the object to forge an appropriate \Q@Point3D@
to store. Note how all the information about what fields sit in
\Q@Point3D@ and \Q@Point2D@ is properly encapsulated in the
\Q@with@ method, and is transparent to the implementer of \Q@Bird@.

@@

\paragraph{Generated boilerplate}
Just to give a feeling of how much mechanical code \mixin is generating, we show the
generated code for the \texttt{Point3D} in Figure~\ref{fig:boilerplate}. Writing such code by hand is error-prone. For
example a distracted programmer may swap the arguments of calls to
\Q@Point3D.of@.  Note how \Q@with-@ methods are automatically refined in their
return type, so that code like:

@@

\subsection{Advanced Multiple Inheritance with Object Interfaces}
Finally, defining \texttt{Pegasus} is as simple as we did in the simplified
(and stateless) version on the right of Figure~\ref{fig:pegasus}.
 Note how even the non-trivial pattern for field type refinement is
transparently composed, and \texttt{Pegasus} has a \Q@Point3D@
\Q@location@.%  This works because \Q@Horse@ do not perform
% any field type refinement, otherwise we may have to choose/create a
% common subtype in order for \texttt{Pegasus} to exists.
